Imperfect slope measurements drive overestimation in geometric cone model of waterbody depth

### 2021-11-10 manuscript/manuscript.Rmd
good looking render of accepted ms

# Optional: line of arbitrary text with additional information.
# Could be used, for example, to mention the bibliographic info in a post-print.
# If not specified, defaults to "This version was compiled on \today"
date_subtitle: > 
  This is the accepted version of the following article: J. Stachelek et al., Inland Waters. In Press.  This article may be used for non-commercial purposes in accordance with Wiley Terms and Conditions for Use of Self-Archived Versions.

@@

where the product of slope and horizontal distance yields an exact geometric depth estimate ($depth_{geometric}$). Cone models of waterbody basins have been used extensively to estimate hypsography in waterbodies where there is no knowledge of volume or mean depth \citep{readSimulating2368Temperate2014, winslowLargescaleModeledContemporary2017}.

@@

As a result of this limited testing, we lack knowledge on both the predictive performance of geometric models, the effect of proxies on depth prediction, and whether depth predictions are more sensitive to measurement errors in the horizontal dimension (i.e. distance to the deepest point of the waterbody) or measurement errors in the vertical dimension (i.e. in-lake slope). Additionally, it is unclear whether model prediction error is related to differences in waterbody type such those with different cross-section shapes (concave versus convex) or those classified as reservoirs versus natural lakes. To address these knowledge gaps, we asked three research questions: (1) How representative is nearshore land slope of in-lake slope; and how representative is the distance to the center of a waterbody compared to the distance to the deepest point of a waterbody? (2) How does the use of proxies for waterbody geometry affect waterbody depth prediction error? (3) How does waterbody cross-section shape (i.e. concave versus convex) and waterbody type (i.e. natural lake vs reservoir) affect depth prediction error? To answer these questions, we extracted maximum depth (hereafter referred to as “observed maximum depth”), in-lake slope, cross-section shape (i.e., concave versus convex), and distance to the deepest point, of approximately 5,000 waterbodies from bathymetric map data. We supplemented these geometry measures with data classifying waterbodies as reservoirs or natural lakes. We used this data to compute geometric depth estimates (Equation \ref{eq1}) and prediction "offsets" to these estimates using the random forest algorithm (Equation \ref{eq3}). Such offsets are model quantities which minimize differences between observed and predicted depth as a function of covariates. In our case, covariates included a variety of waterbody, watershed, and hydrologic subbasin measures that are available for all waterbodies (Table \ref{table1}).

@@

The original data came in a variety of formats including pre-interpolated rasters (Minnesota), contour lines (Nebraska, Michigan, Massachusetts, Kansas, Iowa), contour polygons (New Hampshire, Connecticut), or point depth soundings (Maine). For the Minnesota data, we simply clipped the raster for each waterbody to its outline. For data from the remaining States, we processed each waterbody by converting its original representation to a point layer (if necessary), rasterizing these points, and creating an interpolated bathymetry “surface” using a simple moving window average in the raster R package \citep{hijmansRasterGeographicData2019}. The size of the moving window was adjusted iteratively to ensure that each bathymetry raster contained no missing data.

@@

We used our generated bathymetry surfaces to find the location of the deepest point in the waterbody and we resolved ties by choosing the deepest point that was closest to the center of the waterbody. We used the location of this deepest point to calculate "distance to the deepest point" as the minimum distance to the waterbody shoreline. To account for waterbodies where the centroid does not intersect waterbody bathymetry because it is located within an embedded island or peninsula, we calculated the center of the waterbody not as its centroid but rather by finding the point farthest from the waterbody shoreline (i.e. its “visual distance to waterbody center”). For these calculations, we used the polylabelr R package \citep{larssonPolylabelrFindPole2019}, which interfaces with the Mapbox pole of inaccessibility algorithm \citep{agafonkinJSLibraryFinding2019}. We calculated (maximum) in-lake slope as depth at the deepest point divided by the shortest distance to the deepest point from the shoreline.. We calculated (mean) nearshore land slope for each waterbody by computing the slope within a 100-m buffer using data from a high-resolution digital elevation model (~ 15x15m grain) accessed using the elevatr R package \citep{hollisterElevatrAccessElevation2017}. We explored alternative buffer sizes ranging between 50-1000 m following \citet{sobekPredictingDepthVolume2011} and although 100 m provided the lowest model error, we ultimately found little appreciable effect of varying buffers on model performance. Slope computations proceeded by passing a 3x3 moving window over the 100-m buffer to calculate the slope at each point using Horn's algorithm via the terrain function in the raster R package \citep{hijmansRasterGeographicData2019}. Reported nearshore land slope values are the mean of all points in the buffer. In addition to the aforementioned techniques of calculating in-lake (and nearshore) slopes and distances, we tried 7 alternate techniques which are described in Fig. S5 and Table S2 including measures such as median slope (results not shown).

@@

where sinuous waterbodies have larger values of shoreline development and circular waterbodies have smaller values of shoreline development. Watershed to waterbody area ratio is an approximation of water residence time and is defined as watershed area divided by waterbody area \citep{timmsGeomorphologyLakeBasins2009}.

### Proxy evaluation
We conducted a qualitative assessment of whether or not proxy measures of waterbody geometry (e.g.  nearshore land slope, distance to the center of the waterbody) are representative of their true values (e.g. in-lake slope, distance to the deepest point of the waterbody) by visual inspection (i.e. plotting each proxy measure against its corresponding true value) and by computing coefficients of determination ($R^2$). We further tested proxy measures by examining their effect on waterbody depth prediction error. Our approach involved several steps. In the first step, we computed a geometric estimate of waterbody depth using only geometry information ($depth_{geometric}$, Equation \ref{eq1}). In the second step, we fit a random forest model to predict observed (i.e. true) depth as a function of geometric depth along with several covariates available for all waterbodies (Table \ref{table1}). The purpose of this random forest “offset” modeling was to more rigorously test our expectations regarding prediction error among different formulations of $depth_{geometric}$ and among different waterbody types. Each of these steps were executed iteratively for each combination of true and proxy values of slope and distance (Table 2). We conducted additional sensitivity analysis to examine possible interactions between different proxy measures of waterbody geometry and different subsets of the entire dataset where model data was restricted (i.e. “filtered”) to include only reservoirs, only natural lakes, only convex waterbodies, or only concave waterbodies (Table 2).

\begin{table}
  \caption{Model fit and predictive accuracy metrics (RMSE = root mean square error, R2 = coefficient of determination, MAPE = mean absolute percent error) for all combinations of true (in-lake slope, distance to the deepest point of the waterbody) and proxy (nearshore land slope, distance to waterbody center) metrics.} \label{table2}
  \centering
  % \setlength\tabcolsep{1.5pt} % default value: 6pt
\begin{tabular}{llll}
  \hline
  slope & distance & RMSE & $R^2$\\
  \hline
  true & true & - & -\\
  true & proxy & 4.2 m & 0.73\\
  proxy & true & 6.9 m & 0.26\\
  proxy & proxy & 6.6 m & 0.31\\
  \bottomrule
\end{tabular}
\end{table}

@@

### Random forest models
Prior studies using geometric models to predict waterbody depth include a statistical or machine learning model “layer” or “offset” to boost predictive accuracy \citep{hollisterPredictingMaximumLake2011, yigzawNewGlobalStorage2018}. This procedure involves fitting a statistical or machine learning model to the residuals of an initial geometric model. For our purposes, such offset modeling enabled us to test our expectations that prediction error would be different among different formulations of $depth_{geometric}$ and among different waterbody types. It also facilitated direct comparison against prior models of waterbody depth including those that are non-geometric. We generated an “offset” to geometric depth (sensu \citealt{hollisterPredictingMaximumLake2011}) using the random forest algorithm and the ranger R package \citep{wrightRangerFastImplementation2017} to predict observed maximum depth as a function of covariates including geometric maximum depth (from Equation \ref{eq1}) along with the waterbody elevation, area, perimeter, and ratio/index measures listed in Table 2:

@@

We evaluated the relative “importance” of individual covariates by comparing model performance between 1) models where a given covariate was left untouched versus 2) models where a given covariate was permuted randomly \citep{prasadNewerClassificationRegression2006, wrightRangerFastImplementation2017}. Neither cross-section shape nor reservoir class was used as a covariate in any random forest models. Random forest training and test data was stratified on shape and reservoir class to match that of the overall waterbody population. We used the random forest algorithm because it makes no assumptions about the distribution of model residuals, allows for non-linearity, and is insensitive to interactions (i.e. multicollinearity) among covariates \citep{prasadNewerClassificationRegression2006}.

@@

The use of proxy nearshore land slope had a larger effect on model fit and prediction error than the use of proxy distance to waterbody center (Table 2). More specifically, the true slope (in-lake slope) - proxy distance (distance to the center of the waterbody) model had a better fit ($R^2$ = 0.73) and lower prediction error (RMSE = 4.8m, MAPE = 27%) compared to the proxy slope - true distance model ($R^2$ = 0.31, RMSE = 7.3m, MAPE = 64%). The fit of the proxy slope -proxy distance model ($R^2$ = 0.36, RMSE = 7.1m, MAPE = 61%) was very similar to the proxy slope – true distance model. Predicted depth values for this model were generally underestimates relative to measured depth values (Fig. S9).

Furthermore, analysis of model residuals showed overestimation of waterbody depth for concave waterbodies when models included a proxy slope measure (Fig. \ref{fig4}). We observed similar but smaller overestimation depending on if a waterbody was classified as a reservoir rather than a natural lake (Fig. \ref{fig4}). We found that models restricted to consider only concave lakes had lower error (both in absolute and relative terms) compared to models on other data subsets (e.g. convex lakes, reservoirs, natural lakes, see Table 2). Conversely, we did not observe any notable geographic patterns in model residuals (Fig. S6).

@@

### Effects of proxy measures of waterbody geometry depth prediction error
Models using only proxy variables (Table 2) had prediction error rates (RMSE = 7.1m) of a similar magnitude as that of prior studies (RMSE = 6 - 7.3m) predicting waterbody depth at broad geographic extents \citep{hollisterPredictingMaximumLake2011, messagerEstimatingVolumeAge2016, oliver2016prediction}. When only a single proxy measure was used, there was a difference in model sensitivity depending on if it was a horizontal distance measure or a vertical slope measure. In the case of a true slope and proxy distance combination, models were more accurate (± 4.8m, 27%) than even the most accurate of prior studies \citep{hollisterPredictingMaximumLake2011, messagerEstimatingVolumeAge2016, oliver2016prediction}. Conversely, models using a proxy slope and true distance combination had prediction error rates (± 7.3m, 64%) of a similar magnitude as that of the baseline proxy-proxy model (± 7.1m, 61%). The greater sensitivity of depth predictions to proxy slope measures relative to proxy distance measures may be explained by the fact that proxy slope measures were a more imperfect representation of true in-lake slopes relative to proxy versus true distances. We did not find evidence that the sensitivity of depth predictions to slope was dependent on variations in how these measures were calculated (Table S2). In a general sense, the sensitivity of depth predictions to slope help explain the relatively poor predictive performance of prior non-geometric waterbody depth models given that they rely heavily on waterbody area as a predictor \citep{sobekPredictingDepthVolume2011, messagerEstimatingVolumeAge2016, oliver2016prediction} and both horizontal distance measures and vertical slope measures appear to be decoupled from waterbody area (Fig. S8).

>>>>

# Optional: line of arbitrary text with additional information.
# Could be used, for example, to mention the bibliographic info in a post-print.
# If not specified, defaults to "This version was compiled on \today"
date_subtitle: > 
  This is an Accepted Manuscript of an article to be published by Taylor & Francis in Inland Waters.

@@

\noindent where the product of slope and horizontal distance yields an exact geometric depth estimate ($depth_{geometric}$). Cone models of waterbody basins have been used extensively to estimate hypsography in waterbodies where there is no knowledge of volume or mean depth \citep{readSimulating2368Temperate2014, winslowLargescaleModeledContemporary2017}.

@@

As a result of this limited testing, we lack knowledge on both the predictive performance of geometric models, the effect of proxies on depth prediction, and whether depth predictions are more sensitive to measurement errors in the horizontal dimension (i.e. distance to the deepest point of the waterbody) or measurement errors in the vertical dimension (i.e. in-lake slope). Additionally, it is unclear whether model prediction error is related to differences in waterbody type such those with different cross-section shapes (concave versus convex) or those classified as reservoirs versus natural lakes. To address these knowledge gaps, we asked three research questions:

\begin{enumerate}
\item How representative is nearshore land slope of in-lake slope; and how representative is the distance to the center of a waterbody compared to the distance to the deepest point of a waterbody?
\item How does the use of proxies for waterbody geometry affect waterbody depth prediction error?
\item How does waterbody cross-section shape (i.e. concave versus convex) and waterbody type (i.e. natural lake vs reservoir) affect depth prediction error?
\end{enumerate}

To answer these questions, we extracted maximum depth (hereafter referred to as “observed maximum depth”), in-lake slope, cross-section shape (i.e., concave versus convex), and distance to the deepest point, of approximately 5,000 waterbodies from bathymetric map data. We supplemented these geometry measures with data classifying waterbodies as reservoirs or natural lakes. We used this data to compute geometric depth estimates (Equation \ref{eq1}) and prediction "offsets" to these estimates using the random forest algorithm (Equation \ref{eq3}). Such offsets are model quantities which minimize differences between observed and predicted depth as a function of covariates. In our case, covariates included a variety of waterbody, watershed, and hydrologic subbasin measures that are available for all waterbodies (Table \ref{table1}).

@@

The original data came in a variety of formats including pre-interpolated rasters (Minnesota), contour lines (Nebraska, Michigan, Massachusetts, Kansas, Iowa), contour polygons (New Hampshire, Connecticut), or point depth soundings (Maine). For the Minnesota data, we simply clipped the raster for each waterbody to its outline. For data from the remaining States, we processed each waterbody by converting its original representation to a point layer (if necessary), rasterizing these points, and creating an interpolated bathymetry “surface” using a simple moving window average in the \texttt{raster} R package \citep{hijmansRasterGeographicData2019}. The size of the moving window was adjusted iteratively to ensure that each bathymetry raster contained no missing data.

@@

We used our generated bathymetry surfaces to find the location of the deepest point in the waterbody and we resolved ties by choosing the deepest point that was closest to the center of the waterbody. We used the location of this deepest point to calculate "distance to the deepest point" as the minimum distance to the waterbody shoreline. To account for waterbodies where the centroid does not intersect waterbody bathymetry because it is located within an embedded island or peninsula, we calculated the center of the waterbody not as its centroid but rather by finding the point farthest from the waterbody shoreline (i.e. its “visual distance to waterbody center”). For these calculations, we used the \texttt{polylabelr} R package \citep{larssonPolylabelrFindPole2019}, which interfaces with the Mapbox pole of inaccessibility algorithm \citep{agafonkinJSLibraryFinding2019}. We calculated (maximum) in-lake slope as depth at the deepest point divided by the shortest distance to the deepest point from the shoreline. We calculated (mean) nearshore land slope for each waterbody by computing the slope within a 100-m buffer using data from a high-resolution digital elevation model (~ 15x15m grain) accessed using the \texttt{elevatr} R package \citep{hollisterElevatrAccessElevation2017}. We explored alternative buffer sizes ranging between 50-1000 m following \citet{sobekPredictingDepthVolume2011} and although 100 m provided the lowest model error, we ultimately found little appreciable effect of varying buffers on model performance. Slope computations proceeded by passing a 3x3 moving window over the 100-m buffer to calculate the slope at each point using Horn's algorithm via the terrain function in the \texttt{raster} R package \citep{hijmansRasterGeographicData2019}. Reported nearshore land slope values are the mean of all points in the buffer. In addition to the aforementioned techniques of calculating in-lake (and nearshore) slopes and distances, we tried 7 alternate techniques which are described in Fig. S5 and Table S2 including measures such as median slope (results not shown).

@@

\noindent where sinuous waterbodies have larger values of shoreline development and circular waterbodies have smaller values of shoreline development. Watershed to waterbody area ratio is an approximation of water residence time and is defined as watershed area divided by waterbody area \citep{timmsGeomorphologyLakeBasins2009}.

### Proxy evaluation
We conducted a qualitative assessment of whether or not proxy measures of waterbody geometry (e.g.  nearshore land slope, distance to the center of the waterbody) are representative of their true values (e.g. in-lake slope, distance to the deepest point of the waterbody) by visual inspection (i.e. plotting each proxy measure against its corresponding true value) and by computing coefficients of determination ($R^2$). We further tested proxy measures by examining their effect on waterbody depth prediction error. Our approach involved several steps. In the first step, we computed a geometric estimate of waterbody depth using only geometry information ($depth_{geometric}$, Equation \ref{eq1}). In the second step, we fit a random forest model to predict observed (i.e. true) depth as a function of geometric depth along with several covariates available for all waterbodies (Table \ref{table1}). The purpose of this random forest “offset” modeling was to more rigorously test our expectations regarding prediction error among different formulations of $depth_{geometric}$ and among different waterbody types. Each of these steps were executed iteratively for each combination of true and proxy values of slope and distance (Table \ref{table2}). We conducted additional sensitivity analysis to examine possible interactions between different proxy measures of waterbody geometry and different subsets of the entire dataset where model data was restricted (i.e. “filtered”) to include only reservoirs, only natural lakes, only convex waterbodies, or only concave waterbodies (Table \ref{table2}).

\begin{table}
  \caption{Model fit and predictive accuracy metrics (RMSE = root mean square error, R2 = coefficient of determination, MAPE = mean absolute percent error) for all combinations of true (in-lake slope, distance to the deepest point of the waterbody) and proxy (nearshore land slope, distance to waterbody center) metrics.} \label{table2}
  \centering
  % \setlength\tabcolsep{1.5pt} % default value: 6pt
\begin{tabular}{llllll}
\hline
Filter & Slope & Distance & RMSE & $R^2$ & MAPE\\
\hline
all & true & true &  &  & \\
 & true & proxy & 4.8 m & 0.73 & 27 \%\\
 & proxy & true & 7.3 m & 0.31 & 64 \%\\
 & proxy & proxy & 7.1 m & 0.36 & 61 \%\\
\hline
reservoir & true & true &  &  & \\
 & true & proxy & 5.3 m & 0.66 & 36 \%\\
 & proxy & true & 7 m & 0.40 & 61 \%\\
 & proxy & proxy & 7 m & 0.41 & 60 \%\\
\hline
natural lake & true & true &  &  & \\
 & true & proxy & 4.1 m & 0.74 & 22 \%\\
 & proxy & true & 6.7 m & 0.26 & 68 \%\\
 & proxy & proxy & 6.6 m & 0.29 & 64 \%\\
\hline
convex & true & true &  &  & \\
 & true & proxy & 4.7 m & 0.74 & 30 \%\\
 & proxy & true & 7.2 m & 0.34 & 59 \%\\
 & proxy & proxy & 6.9 m & 0.39 & 58 \%\\
\hline
concave & true & true &  &  & \\
 & true & proxy & 1.6 m & 0.78 & 20 \%\\
 & proxy & true & 3.1 m & 0.14 & 46 \%\\
 & proxy & proxy & 3 m & 0.17 & 45 \%\\
\bottomrule
\end{tabular}
\end{table}

@@

### Random forest models
Prior studies using geometric models to predict waterbody depth include a statistical or machine learning model “layer” or “offset” to boost predictive accuracy \citep{hollisterPredictingMaximumLake2011, yigzawNewGlobalStorage2018}. This procedure involves fitting a statistical or machine learning model to the residuals of an initial geometric model. For our purposes, such offset modeling enabled us to test our expectations that prediction error would be different among different formulations of $depth_{geometric}$ and among different waterbody types. It also facilitated direct comparison against prior models of waterbody depth including those that are non-geometric. We generated an “offset” to geometric depth (sensu \citealt{hollisterPredictingMaximumLake2011}) using the random forest algorithm and the \texttt{ranger} R package \citep{wrightRangerFastImplementation2017} to predict observed maximum depth as a function of covariates including geometric maximum depth (from Equation \ref{eq1}) along with the waterbody elevation, area, perimeter, and ratio/index measures listed in Table \ref{table2}:

@@

\noindent We evaluated the relative “importance” of individual covariates by comparing model performance between 1) models where a given covariate was left untouched versus 2) models where a given covariate was permuted randomly \citep{prasadNewerClassificationRegression2006, wrightRangerFastImplementation2017}. Neither cross-section shape nor reservoir class was used as a covariate in any random forest models. Random forest training and test data was stratified on shape and reservoir class to match that of the overall waterbody population. We used the random forest algorithm because it makes no assumptions about the distribution of model residuals, allows for non-linearity, and is insensitive to interactions (i.e. multicollinearity) among covariates \citep{prasadNewerClassificationRegression2006}.

@@

The use of proxy nearshore land slope had a larger effect on model fit and prediction error than the use of proxy distance to waterbody center (Table \ref{table2}). More specifically, the true slope (in-lake slope) - proxy distance (distance to the center of the waterbody) model had a better fit ($R^2$ = 0.73) and lower prediction error (RMSE = 4.8m, MAPE = 27%) compared to the proxy slope - true distance model ($R^2$ = 0.31, RMSE = 7.3m, MAPE = 64%). The fit of the proxy slope -proxy distance model ($R^2$ = 0.36, RMSE = 7.1m, MAPE = 61%) was very similar to the proxy slope – true distance model. Predicted depth values for this model were generally underestimates relative to measured depth values (Fig. S9).

Furthermore, analysis of model residuals showed overestimation of waterbody depth for concave waterbodies when models included a proxy slope measure (Fig. \ref{fig4}). We observed similar but smaller overestimation depending on if a waterbody was classified as a reservoir rather than a natural lake (Fig. \ref{fig4}). We found that models restricted to consider only concave lakes had lower error (both in absolute and relative terms) compared to models on other data subsets (e.g. convex lakes, reservoirs, natural lakes, see Table \ref{table2}). Conversely, we did not observe any notable geographic patterns in model residuals (Fig. S6).

@@

### Effects of proxy measures of waterbody geometry depth prediction error
Models using only proxy variables (Table \ref{table2}) had prediction error rates (RMSE = 7.1m) of a similar magnitude as that of prior studies (RMSE = 6 - 7.3m) predicting waterbody depth at broad geographic extents \citep{hollisterPredictingMaximumLake2011, messagerEstimatingVolumeAge2016, oliver2016prediction}. When only a single proxy measure was used, there was a difference in model sensitivity depending on if it was a horizontal distance measure or a vertical slope measure. In the case of a true slope and proxy distance combination, models were more accurate (± 4.8m, 27%) than even the most accurate of prior studies \citep{hollisterPredictingMaximumLake2011, messagerEstimatingVolumeAge2016, oliver2016prediction}. Conversely, models using a proxy slope and true distance combination had prediction error rates (± 7.3m, 64%) of a similar magnitude as that of the baseline proxy-proxy model (± 7.1m, 61%). The greater sensitivity of depth predictions to proxy slope measures relative to proxy distance measures may be explained by the fact that proxy slope measures were a more imperfect representation of true in-lake slopes relative to proxy versus true distances. We did not find evidence that the sensitivity of depth predictions to slope was dependent on variations in how these measures were calculated (Table S2). In a general sense, the sensitivity of depth predictions to slope help explain the relatively poor predictive performance of prior non-geometric waterbody depth models given that they rely heavily on waterbody area as a predictor \citep{sobekPredictingDepthVolume2011, messagerEstimatingVolumeAge2016, oliver2016prediction} and both horizontal distance measures and vertical slope measures appear to be decoupled from waterbody area (Fig. S8).
